---
title: "health-food"
---

This project investigates physical measurements and health metrics from the National Health and Nutrition Examination Survey (NHANES), provided by the `primer.data` R package.

```{r}
#| label: setup
#| include: false

library(tidyverse)
```

```{r}
#| label: obesity-time-plot
#| fig-width: 9
#| fig-height: 5
#| fig-align: center

obesity_trends <- tibble(
  year = c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020, 2023),
  rate = c(0.305, 0.305, 0.322, 0.343, 0.337, 0.357, 0.349, 0.377, 0.396, 0.424, 0.419, 0.403)
)

obesity_trends |>
  ggplot(aes(x = year, y = rate)) +
  geom_line(color = "#2c7bb6", linewidth = 1.2) +
  geom_point(color = "#2c7bb6", size = 3) +
  geom_text(aes(label = scales::percent(rate, accuracy = 0.1)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.25, 0.48),
    breaks = seq(0.25, 0.45, by = 0.05)
  ) +
  scale_x_continuous(breaks = c(2000, 2004, 2008, 2012, 2016, 2020, 2023)) +
  labs(
    title = "US Adult Obesity Prevalence Over Time (1999–2023)",
    subtitle = "Age-adjusted percentage of adults (aged 20+) with BMI ≥ 30",
    x = "Survey Year",
    y = "Obesity Rate",
    caption = "Source: CDC / National Center for Health Statistics (NHANES)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

## Dietary Trends: Ultra-Processed vs. Minimally Processed Foods

Research from NHANES dietary recall surveys tracks the proportion of daily calories coming from ultra-processed foods compared to whole and minimally processed foods among Americans.

```{r}
#| label: food-processing-plot
#| fig-width: 9
#| fig-height: 5.5
#| fig-align: center

food_trends <- bind_rows(
  tibble(
    year = c(2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2023),
    pct_calories = c(0.535, 0.543, 0.554, 0.558, 0.561, 0.564, 0.569, 0.567, 0.570, 0.530),
    category = "Ultra-Processed Foods"
  ),
  tibble(
    year = c(2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2023),
    pct_calories = c(0.327, 0.318, 0.305, 0.300, 0.295, 0.291, 0.283, 0.281, 0.274, 0.305),
    category = "Whole & Minimally Processed Foods"
  )
)

food_trends |>
  ggplot(aes(x = year, y = pct_calories, color = category)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(
    aes(label = scales::percent(pct_calories, accuracy = 0.1)),
    vjust = -1,
    size = 3.5,
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = c(
      "Ultra-Processed Foods" = "#e6550d",
      "Whole & Minimally Processed Foods" = "#2ca25f"
    )
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.24, 0.62),
    breaks = seq(0.25, 0.60, by = 0.05)
  ) +
  scale_x_continuous(breaks = c(2002, 2006, 2010, 2014, 2018, 2023)) +
  labs(
    title = "US Caloric Share: Ultra-Processed vs. Minimally Processed (2001–2023)",
    subtitle = "Percentage of total daily calories among US adults by processing category",
    x = "Survey Period Ending Year",
    y = "Share of Daily Calories",
    color = "Food Category",
    caption = "Source: NHANES dietary intake data / Juul et al. (2022) & NCHS Data Brief No. 536 (2025)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

- **Opposing Trends:** As ultra-processed foods rose from 53.5% in 2001–2002 to 57.0% in 2017–2018, whole and minimally processed foods fell from 32.7% to 27.4%.
- **Recent Shift:** In the August 2021–August 2023 survey cycle, ultra-processed food intake dropped to 53.0%, while whole and minimally processed calories rose back to approximately 30.5%.

