---
title: "health-food"
---

This project investigates physical measurements and health metrics from the National Health and Nutrition Examination Survey (NHANES), provided by the `primer.data` R package.

```{r}
#| label: setup
#| include: false

library(tidyverse)
```

```{r}
#| label: obesity-time-plot
#| fig-width: 9
#| fig-height: 5
#| fig-align: center

obesity_trends <- tibble(
  year = c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020),
  rate = c(0.305, 0.305, 0.322, 0.343, 0.337, 0.357, 0.349, 0.377, 0.396, 0.424, 0.419)
)

obesity_trends |>
  ggplot(aes(x = year, y = rate)) +
  geom_line(color = "#2c7bb6", linewidth = 1.2) +
  geom_point(color = "#2c7bb6", size = 3) +
  geom_text(aes(label = scales::percent(rate, accuracy = 0.1)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.25, 0.48),
    breaks = seq(0.25, 0.45, by = 0.05)
  ) +
  scale_x_continuous(breaks = seq(2000, 2020, by = 4)) +
  labs(
    title = "US Adult Obesity Prevalence Over Time (1999–2020)",
    subtitle = "Age-adjusted percentage of adults (aged 20+) with BMI ≥ 30",
    x = "Survey Year",
    y = "Obesity Rate",
    caption = "Source: CDC / National Center for Health Statistics (NHANES)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

## Ultra-Processed Food Consumption Over Time

Research from NHANES dietary recall surveys tracks the proportion of daily calories coming from ultra-processed foods (such as packaged snacks, sugary beverages, and ready-to-eat meals) among Americans.

```{r}
#| label: upf-time-plot
#| fig-width: 9
#| fig-height: 5
#| fig-align: center

upf_trends <- tibble(
  period = c("2001–2002", "2003–2004", "2005–2006", "2007–2008", 
             "2009–2010", "2011–2012", "2013–2014", "2015–2016", "2017–2018"),
  year = c(2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018),
  pct_calories = c(0.535, 0.543, 0.554, 0.558, 0.561, 0.564, 0.569, 0.567, 0.570)
)

upf_trends |>
  ggplot(aes(x = year, y = pct_calories)) +
  geom_line(color = "#e6550d", linewidth = 1.2) +
  geom_point(color = "#e6550d", size = 3) +
  geom_text(aes(label = scales::percent(pct_calories, accuracy = 0.1)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.48, 0.62),
    breaks = seq(0.48, 0.62, by = 0.02)
  ) +
  scale_x_continuous(breaks = seq(2002, 2018, by = 2)) +
  labs(
    title = "Ultra-Processed Food Consumption Over Time (2001–2018)",
    subtitle = "Percentage of total daily calories from ultra-processed food among US adults",
    x = "Survey Period Ending Year",
    y = "Share of Daily Calories",
    caption = "Source: NHANES dietary intake data / Juul et al. (Am J Clin Nutr, 2022)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

- **Majority of Calories:** Ultra-processed foods contribute over 57% of daily calories consumed by US adults.
- **Steady Rise:** Between 2001–2002 and 2017–2018, the share of daily energy from ultra-processed food grew by 3.5 percentage points.

