---
title: "NHANES Health Insights"
subtitle: "Exploring obesity rates across age and sex"
---

The **National Health and Nutrition Examination Survey (NHANES)** assesses the health and nutritional status of adults and children across the United States.

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(primer.data)
```

```{r}
#| label: obesity-plot
#| fig-width: 10
#| fig-height: 5.5
#| fig-align: center

nhanes |>
  filter(age >= 18) |>
  drop_na(age, bmi, sex) |>
  mutate(obese = if_else(bmi >= 30, 1, 0)) |>
  group_by(age, sex) |>
  summarize(obesity_rate = mean(obese), .groups = "drop") |>
  ggplot(aes(x = age, y = obesity_rate, color = sex)) +
  geom_point(alpha = 0.4, size = 1.8) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  facet_wrap(~ sex) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_manual(values = c("Female" = "#d95f02", "Male" = "#1b9e77")) +
  labs(
    title = "Obesity Rates Across Age",
    subtitle = "Percentage of adults with BMI ≥ 30 by age and sex",
    x = "Age (years)",
    y = "Obesity Rate",
    caption = "Source: National Health and Nutrition Examination Survey (primer.data::nhanes)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

### Key Observations

- **Young Adulthood:** Obesity rates start around 20–30% in early adulthood for both males and females.
- **Middle Adulthood:** Obesity prevalence rises through midlife, peaking near 40–45% between ages 45 and 65.
- **Older Age:** Obesity rates decline in later decades, likely reflecting both metabolic shifts and survivorship effects.


