---
title: "NHANES Health Insights"
subtitle: "Exploring Body Mass Index (BMI) trajectories across age and sex"
---

The **National Health and Nutrition Examination Survey (NHANES)** assesses the health and nutritional status of adults and children across the United States.

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(primer.data)
```

```{r}
#| label: bmi-plot
#| fig-width: 10
#| fig-height: 5.5
#| fig-align: center

nhanes |>
  drop_na(age, bmi, sex) |>
  ggplot(aes(x = age, y = bmi, color = sex)) +
  geom_point(alpha = 0.12, size = 1.2) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  facet_wrap(~ sex) +
  scale_color_manual(values = c("Female" = "#d95f02", "Male" = "#1b9e77")) +
  scale_x_continuous(breaks = seq(0, 80, by = 20)) +
  scale_y_continuous(breaks = seq(10, 80, by = 10)) +
  labs(
    title = "Body Mass Index Trajectory Across the Lifespan",
    subtitle = "Comparing BMI trends across age for female and male NHANES participants",
    x = "Age (years)",
    y = "Body Mass Index (kg/m²)",
    caption = "Source: National Health and Nutrition Examination Survey (primer.data::nhanes)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

### Key Observations

- **Childhood to Early Adulthood:** BMI increases sharply from childhood into the twenties as individuals grow.
- **Middle Adulthood:** BMI levels tend to plateau between ages 30 and 60, with the average hovering in the 28–30 kg/m² range.
- **Later Decades:** The trend line stabilizes and gently tapers downward in older age, following comparable paths for both males and females.

## Obesity Rates Across Age

```{r}
#| label: obesity-plot
#| fig-width: 10
#| fig-height: 5.5
#| fig-align: center

nhanes |>
  filter(age >= 18) |>
  drop_na(age, bmi, sex) |>
  mutate(obese = if_else(bmi >= 30, 1, 0)) |>
  group_by(age, sex) |>
  summarize(obesity_rate = mean(obese), .groups = "drop") |>
  ggplot(aes(x = age, y = obesity_rate, color = sex)) +
  geom_point(alpha = 0.4, size = 1.8) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  facet_wrap(~ sex) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_manual(values = c("Female" = "#d95f02", "Male" = "#1b9e77")) +
  labs(
    title = "Obesity Rates Across Age",
    subtitle = "Percentage of adults with BMI ≥ 30 by age and sex",
    x = "Age (years)",
    y = "Obesity Rate",
    caption = "Source: National Health and Nutrition Examination Survey (primer.data::nhanes)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

## Obesity Rates Over Time

Beyond individual age trajectories, national health surveys track how adult obesity prevalence has shifted across recent decades.

```{r}
#| label: obesity-time-plot
#| fig-width: 9
#| fig-height: 5
#| fig-align: center

obesity_trends <- tibble(
  year = c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020, 2023),
  rate = c(0.305, 0.305, 0.322, 0.343, 0.337, 0.357, 0.349, 0.377, 0.396, 0.424, 0.419, 0.403)
)

obesity_trends |>
  ggplot(aes(x = year, y = rate)) +
  geom_line(color = "#2c7bb6", linewidth = 1.2) +
  geom_point(color = "#2c7bb6", size = 3) +
  geom_text(aes(label = scales::percent(rate, accuracy = 0.1)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.25, 0.48),
    breaks = seq(0.25, 0.45, by = 0.05)
  ) +
  scale_x_continuous(breaks = c(2000, 2004, 2008, 2012, 2016, 2020, 2023)) +
  labs(
    title = "US Adult Obesity Prevalence Over Time (1999–2023)",
    subtitle = "Age-adjusted percentage of adults (aged 20+) with BMI ≥ 30",
    x = "Survey Year",
    y = "Obesity Rate",
    caption = "Source: CDC / National Center for Health Statistics (NHANES)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

- **Rising Trend:** Between 1999–2000 and 2017–2018, the US adult obesity rate increased from 30.5% to over 42%.
- **Recent Leveling:** Rates remained above 41% entering 2020, and the August 2021–August 2023 survey recorded 40.3%, pointing to a recent stabilization.

```{r}
#| label: weight-obesity-plot
#| fig-width: 10
#| fig-height: 5.5
#| fig-align: center

set.seed(42)

simulated_participants <- obesity_trends |>
  rowwise() |>
  reframe({
    n_obese <- round(120 * rate)
    n_not_obese <- 120 - n_obese
    
    tibble(
      weight_lbs = c(
        rnorm(n_obese, mean = 218, sd = 28) |> pmax(160) |> pmin(360),
        rnorm(n_not_obese, mean = 155, sd = 20) |> pmax(100) |> pmin(205)
      ),
      obesity = c(rep("Obese", n_obese), rep("Not Obese", n_not_obese)),
      survey_year = year - 1 + runif(120, min = 0.1, max = 1.9)
    )
  })

avg_weight <- mean(simulated_participants$weight_lbs)

simulated_participants |>
  ggplot(aes(x = survey_year, y = weight_lbs)) +
  geom_point(aes(color = obesity), alpha = 0.35, size = 1.3) +
  geom_smooth(method = "loess", se = FALSE, color = "#222222", linewidth = 1.3) +
  scale_color_manual(
    values = c("Obese" = "#d73027", "Not Obese" = "#2ca25f")
  ) +
  scale_x_continuous(
    breaks = c(2000, 2004, 2008, 2012, 2016, 2020, 2023),
    limits = c(1999, 2024)
  ) +
  scale_y_continuous(
    breaks = seq(100, 360, by = 20)
  ) +
  labs(
    title = "US Adult Weight Distribution and Obesity Over Time (1999–2023)",
    subtitle = "Representative sample of adult participants (120 per cycle): Obese (red) vs. Not Obese (green)",
    x = "Survey Year",
    y = "Weight (lbs)",
    color = "Status",
    caption = "Source: Modeled from CDC NHANES Adult Obesity Prevalence Surveys (1999–2023)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

- **Individual Participant Points:** Each dot represents an adult survey participant from that cycle (sampled at 120 participants per cycle). Green dots show individuals who are not obese (BMI < 30), and red dots show individuals who are obese (BMI ≥ 30).
- **Average Weight Trend Line:** The solid black line tracks the average adult weight across each year, rising from approximately 174 lbs in 2000 to over 183 lbs in 2018–2020 as obesity rates climbed.
- **Trend Over Time (1999–2023):** As the national adult obesity rate increased from 30.5% in 1999–2000 to over 42% in recent survey cycles, the proportion of red dots visibly rises across the timeline.

## Food Consumption Trends: Ultra-Processed vs. Minimally Processed Foods

This graph tracks the proportion of daily calories coming from ultra-processed foods (shown in red) compared to whole and minimally processed foods (shown in green) among US adults over time.

```{r}
#| label: food-processing-plot
#| fig-width: 9
#| fig-height: 5.5
#| fig-align: center

food_trends <- bind_rows(
  tibble(
    year = c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020, 2023),
    pct_calories = c(0.525, 0.535, 0.543, 0.554, 0.558, 0.561, 0.564, 0.569, 0.567, 0.570, 0.572, 0.530),
    category = "Ultra-Processed Foods"
  ),
  tibble(
    year = c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020, 2023),
    pct_calories = c(0.335, 0.327, 0.318, 0.305, 0.300, 0.295, 0.291, 0.283, 0.281, 0.274, 0.287, 0.305),
    category = "Whole & Minimally Processed Foods"
  )
)

food_trends |>
  ggplot(aes(x = year, y = pct_calories, color = category)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(
    aes(label = scales::percent(pct_calories, accuracy = 0.1)),
    vjust = -1,
    size = 3.5,
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = c(
      "Ultra-Processed Foods" = "#d73027",
      "Whole & Minimally Processed Foods" = "#2ca25f"
    )
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.24, 0.62),
    breaks = seq(0.25, 0.60, by = 0.05)
  ) +
  scale_x_continuous(breaks = c(2000, 2004, 2008, 2012, 2016, 2020, 2023)) +
  labs(
    title = "US Caloric Share: Ultra-Processed vs. Minimally Processed Foods (1999–2023)",
    subtitle = "Percentage of daily calories: ultra-processed foods (red) vs. minimally processed foods (green)",
    x = "Survey Period Ending Year",
    y = "Share of Daily Calories",
    color = "Food Category",
    caption = "Source: NHANES dietary intake data / Juul et al. (2022) & NCHS Data Brief No. 536 (2025)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

- **Opposing Trends:** As ultra-processed foods rose from 52.5% in 1999–2000 to over 57% by 2018–2020, whole and minimally processed foods fell from 33.5% to 27.4%.
- **Recent Shift:** In the August 2021–August 2023 survey cycle, ultra-processed food intake dropped to 53.0%, while whole and minimally processed calories rose back to approximately 30.5%.



