---
title: "NHANES Health Insights"
subtitle: "Exploring Body Mass Index (BMI) trajectories across age and sex"
---

The **National Health and Nutrition Examination Survey (NHANES)** assesses the health and nutritional status of adults and children across the United States.

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(primer.data)
```

```{r}
#| label: bmi-plot
#| fig-width: 10
#| fig-height: 5.5
#| fig-align: center

nhanes |>
  drop_na(age, bmi, sex) |>
  ggplot(aes(x = age, y = bmi, color = sex)) +
  geom_point(alpha = 0.12, size = 1.2) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  facet_wrap(~ sex) +
  scale_color_manual(values = c("Female" = "#d95f02", "Male" = "#1b9e77")) +
  scale_x_continuous(breaks = seq(0, 80, by = 20)) +
  scale_y_continuous(breaks = seq(10, 80, by = 10)) +
  labs(
    title = "Body Mass Index Trajectory Across the Lifespan",
    subtitle = "Comparing BMI trends across age for female and male NHANES participants",
    x = "Age (years)",
    y = "Body Mass Index (kg/m²)",
    caption = "Source: National Health and Nutrition Examination Survey (primer.data::nhanes)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

### Key Observations

- **Childhood to Early Adulthood:** BMI increases sharply from childhood into the twenties as individuals grow.
- **Middle Adulthood:** BMI levels tend to plateau between ages 30 and 60, with the average hovering in the 28–30 kg/m² range.
- **Later Decades:** The trend line stabilizes and gently tapers downward in older age, following comparable paths for both males and females.

## Obesity Rates Over Time

Beyond individual age trajectories, national health surveys track how adult obesity prevalence has shifted across recent decades.

```{r}
#| label: obesity-time-plot
#| fig-width: 9
#| fig-height: 5
#| fig-align: center

obesity_trends <- tibble(
  year = c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020),
  rate = c(0.305, 0.305, 0.322, 0.343, 0.337, 0.357, 0.349, 0.377, 0.396, 0.424, 0.419)
)

obesity_trends |>
  ggplot(aes(x = year, y = rate)) +
  geom_line(color = "#2c7bb6", linewidth = 1.2) +
  geom_point(color = "#2c7bb6", size = 3) +
  geom_text(aes(label = scales::percent(rate, accuracy = 0.1)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.25, 0.48),
    breaks = seq(0.25, 0.45, by = 0.05)
  ) +
  scale_x_continuous(breaks = seq(2000, 2020, by = 4)) +
  labs(
    title = "US Adult Obesity Prevalence Over Time (1999–2020)",
    subtitle = "Age-adjusted percentage of adults (aged 20+) with BMI ≥ 30",
    x = "Survey Year",
    y = "Obesity Rate",
    caption = "Source: CDC / National Center for Health Statistics (NHANES)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

- **Rising Trend:** Between 1999–2000 and 2017–2018, the US adult obesity rate increased from 30.5% to over 42%.
- **Recent Leveling:** Rates remained above 41% entering 2020, highlighting continuing public health challenges.



