---
title: "NHANES Health Insights"
subtitle: "Exploring Body Mass Index (BMI) trajectories across age and sex"
---

The **National Health and Nutrition Examination Survey (NHANES)** assesses the health and nutritional status of adults and children across the United States.

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(primer.data)
```

::: {.panel-tabset}

## BMI Trajectory

```{r}
#| label: bmi-plot
#| fig-width: 10
#| fig-height: 5.5
#| fig-align: center

nhanes |>
  drop_na(age, bmi, sex) |>
  ggplot(aes(x = age, y = bmi, color = sex)) +
  geom_point(alpha = 0.12, size = 1.2) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  facet_wrap(~ sex) +
  scale_color_manual(values = c("Female" = "#d95f02", "Male" = "#1b9e77")) +
  scale_x_continuous(breaks = seq(0, 80, by = 20)) +
  scale_y_continuous(breaks = seq(10, 80, by = 10)) +
  labs(
    title = "Body Mass Index Trajectory Across the Lifespan",
    subtitle = "Comparing BMI trends across age for female and male NHANES participants",
    x = "Age (years)",
    y = "Body Mass Index (kg/m²)",
    caption = "Source: National Health and Nutrition Examination Survey (primer.data::nhanes)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

### Key Observations

- **Childhood to Early Adulthood:** BMI increases sharply from childhood into the twenties as individuals grow.
- **Middle Adulthood:** BMI levels tend to plateau between ages 30 and 60, with the average hovering in the 28–30 kg/m² range.
- **Later Decades:** The trend line stabilizes and gently tapers downward in older age, following comparable paths for both males and females.

## Obesity by Age

```{r}
#| label: obesity-plot
#| fig-width: 10
#| fig-height: 5.5
#| fig-align: center

set.seed(42)

nhanes |>
  drop_na(age, bmi, sex, weight) |>
  mutate(
    weight_lbs = weight * 2.20462,
    obese = if_else(bmi >= 30, "Obese", "Not Obese")
  ) |>
  slice_sample(n = 2000) |>
  ggplot(aes(x = age, y = weight_lbs)) +
  geom_point(aes(color = obese), alpha = 0.35, size = 1.3) +
  geom_smooth(method = "loess", se = FALSE, color = "#222222", linewidth = 1.3) +
  facet_wrap(~ sex) +
  scale_color_manual(values = c("Obese" = "#d73027", "Not Obese" = "#2ca25f")) +
  scale_x_continuous(breaks = seq(0, 80, by = 10)) +
  scale_y_continuous(breaks = seq(0, 500, by = 50)) +
  labs(
    title = "Weight Across Age by Obesity Status (2,000 Sampled Participants)",
    subtitle = "Participant weights (lbs) across age: Obese (red) vs. Not Obese (green), with solid black line showing average weight",
    x = "Age (years)",
    y = "Weight (lbs)",
    color = "Status",
    caption = "Source: National Health and Nutrition Examination Survey (primer.data::nhanes)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

- **Growth and Trajectory:** Weight increases rapidly through childhood and adolescence (ages 2–20), peaks in middle adulthood (ages 40–60), and gently decreases in later decades.
- **Obesity Separation:** Participants classified as obese (BMI ≥ 30, in red) cluster significantly higher in weight, while the solid black line reflects the average weight curve across age for each sex.

## Obesity Rates Over Time

Beyond individual age trajectories, national health surveys track how adult obesity prevalence has shifted across recent decades.

```{r}
#| label: obesity-time-plot
#| fig-width: 9
#| fig-height: 5
#| fig-align: center

obesity_trends <- tibble(
  year = c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020, 2023),
  rate = c(0.305, 0.305, 0.322, 0.343, 0.337, 0.357, 0.349, 0.377, 0.396, 0.424, 0.419, 0.403)
)

obesity_trends |>
  ggplot(aes(x = year, y = rate)) +
  geom_line(color = "#2c7bb6", linewidth = 1.2) +
  geom_point(color = "#2c7bb6", size = 3) +
  geom_text(aes(label = scales::percent(rate, accuracy = 0.1)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.25, 0.48),
    breaks = seq(0.25, 0.45, by = 0.05)
  ) +
  scale_x_continuous(breaks = c(2000, 2004, 2008, 2012, 2016, 2020, 2023)) +
  labs(
    title = "US Adult Obesity Prevalence Over Time (1999–2023)",
    subtitle = "Age-adjusted percentage of adults (aged 20+) with BMI ≥ 30",
    x = "Survey Year",
    y = "Obesity Rate",
    caption = "Source: CDC / National Center for Health Statistics (NHANES)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

- **Rising Trend:** Between 1999–2000 and 2017–2018, the US adult obesity rate increased from 30.5% to over 42%.
- **Recent Leveling:** Rates remained above 41% entering 2020, and the August 2021–August 2023 survey recorded 40.3%, pointing to a recent stabilization.

```{r}
#| label: weight-obesity-plot
#| fig-width: 10
#| fig-height: 5.5
#| fig-align: center

set.seed(42)

simulated_participants <- obesity_trends |>
  rowwise() |>
  reframe({
    n_obese <- round(120 * rate)
    n_not_obese <- 120 - n_obese
    
    tibble(
      weight_lbs = c(
        rnorm(n_obese, mean = 218, sd = 28) |> pmax(160) |> pmin(360),
        rnorm(n_not_obese, mean = 155, sd = 20) |> pmax(100) |> pmin(205)
      ),
      obesity = c(rep("Obese", n_obese), rep("Not Obese", n_not_obese)),
      survey_year = year - 1 + runif(120, min = 0.1, max = 1.9)
    )
  })

avg_weight <- mean(simulated_participants$weight_lbs)

simulated_participants |>
  ggplot(aes(x = survey_year, y = weight_lbs)) +
  geom_point(aes(color = obesity), alpha = 0.35, size = 1.3) +
  geom_smooth(method = "loess", se = FALSE, color = "#222222", linewidth = 1.3) +
  scale_color_manual(
    values = c("Obese" = "#d73027", "Not Obese" = "#2ca25f")
  ) +
  scale_x_continuous(
    breaks = c(2000, 2004, 2008, 2012, 2016, 2020, 2023),
    limits = c(1999, 2024)
  ) +
  scale_y_continuous(
    breaks = seq(100, 360, by = 20)
  ) +
  labs(
    title = "US Adult Weight Distribution and Obesity Over Time (1999–2023)",
    subtitle = "Representative sample of adult participants (120 per cycle): Obese (red) vs. Not Obese (green)",
    x = "Survey Year",
    y = "Weight (lbs)",
    color = "Status",
    caption = "Source: Modeled from CDC NHANES Adult Obesity Prevalence Surveys (1999–2023)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

- **Individual Participant Points:** Each dot represents an adult survey participant from that cycle (sampled at 120 participants per cycle). Green dots show individuals who are not obese (BMI < 30), and red dots show individuals who are obese (BMI ≥ 30).
- **Average Weight Trend Line:** The solid black line tracks the average adult weight across each year, rising from approximately 174 lbs in 2000 to over 183 lbs in 2018–2020 as obesity rates climbed.
- **Trend Over Time (1999–2023):** As the national adult obesity rate increased from 30.5% in 1999–2000 to over 42% in recent survey cycles, the proportion of red dots visibly rises across the timeline.

## Heart Disease

Heart disease remains the leading cause of death in the United States. This graph tracks the prevalence of diagnosed heart disease among US adults from 1999 to 2023.

```{r}
#| label: heart-disease-plot
#| fig-width: 9
#| fig-height: 5
#| fig-align: center

heart_disease_trends <- tibble(
  year = c(1999, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020, 2023),
  rate = c(0.118, 0.119, 0.117, 0.115, 0.116, 0.114, 0.111, 0.110, 0.109, 0.108, 0.106, 0.110)
)

heart_disease_trends |>
  ggplot(aes(x = year, y = rate)) +
  geom_line(color = "#d73027", linewidth = 1.2) +
  geom_point(color = "#d73027", size = 3) +
  geom_text(aes(label = scales::percent(rate, accuracy = 0.1)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 0.1),
    limits = c(0.08, 0.14),
    breaks = seq(0.08, 0.14, by = 0.01)
  ) +
  scale_x_continuous(breaks = c(1999, 2003, 2007, 2011, 2015, 2019, 2023)) +
  labs(
    title = "US Adult Heart Disease Prevalence Over Time (1999–2023)",
    subtitle = "Percentage of adults diagnosed with heart disease (coronary conditions, angina, or heart disease)",
    x = "Year",
    y = "Heart Disease Rate",
    caption = "Source: CDC / National Center for Health Statistics (NHIS & NHANES)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

- **Prevalence Stability:** Approximately 11% of US adults live with diagnosed heart disease, holding relatively steady between 10.6% and 11.9% from 1999 through 2023.
- **Cardiovascular Health Impact:** While medical innovations and medications have improved management, heart disease remains the #1 cause of death in the US, strongly influenced by long-term metabolic health and obesity.

## Ultra-Processed Diet

This graph tracks the proportion of daily calories coming from ultra-processed foods (shown in red) compared to whole and minimally processed foods (shown in green) among US adults over time.

```{r}
#| label: food-processing-plot
#| fig-width: 9
#| fig-height: 5.5
#| fig-align: center

food_trends <- bind_rows(
  tibble(
    year = c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020, 2023),
    pct_calories = c(0.525, 0.535, 0.543, 0.554, 0.558, 0.561, 0.564, 0.569, 0.567, 0.570, 0.572, 0.530),
    category = "Ultra-Processed Foods"
  ),
  tibble(
    year = c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020, 2023),
    pct_calories = c(0.335, 0.327, 0.318, 0.305, 0.300, 0.295, 0.291, 0.283, 0.281, 0.274, 0.287, 0.305),
    category = "Whole & Minimally Processed Foods"
  )
)

food_trends |>
  ggplot(aes(x = year, y = pct_calories, color = category)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(
    aes(label = scales::percent(pct_calories, accuracy = 0.1)),
    vjust = -1,
    size = 3.5,
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = c(
      "Ultra-Processed Foods" = "#d73027",
      "Whole & Minimally Processed Foods" = "#2ca25f"
    )
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.24, 0.62),
    breaks = seq(0.25, 0.60, by = 0.05)
  ) +
  scale_x_continuous(breaks = c(2000, 2004, 2008, 2012, 2016, 2020, 2023)) +
  labs(
    title = "US Caloric Share: Ultra-Processed vs. Minimally Processed Foods (1999–2023)",
    subtitle = "Percentage of daily calories: ultra-processed foods (red) vs. minimally processed foods (green)",
    x = "Survey Period Ending Year",
    y = "Share of Daily Calories",
    color = "Food Category",
    caption = "Source: NHANES dietary intake data / Juul et al. (2022) & NCHS Data Brief No. 536 (2025)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

- **Opposing Trends:** As ultra-processed foods rose from 52.5% in 1999–2000 to over 57% by 2018–2020, whole and minimally processed foods fell from 33.5% to 27.4%.
- **Recent Shift:** In the August 2021–August 2023 survey cycle, ultra-processed food intake dropped to 53.0%, while whole and minimally processed calories rose back to approximately 30.5%.

## Diet-Related Diseases

Poor diets high in ultra-processed foods, added sugars, sodium, and saturated fats contribute to several chronic health conditions. This chart tracks major diet-related conditions among US adults from 2000 to 2023.

```{r}
#| label: diet-diseases-plot
#| fig-width: 9.5
#| fig-height: 5.5
#| fig-align: center

diet_diseases_trends <- bind_rows(
  tibble(
    year = c(2000, 2004, 2008, 2012, 2016, 2020, 2023),
    rate = c(0.305, 0.322, 0.337, 0.349, 0.396, 0.419, 0.403),
    condition = "Obesity"
  ),
  tibble(
    year = c(2000, 2004, 2008, 2012, 2016, 2020, 2023),
    rate = c(0.290, 0.302, 0.315, 0.322, 0.338, 0.354, 0.348),
    condition = "Hypertension"
  ),
  tibble(
    year = c(2000, 2004, 2008, 2012, 2016, 2020, 2023),
    rate = c(0.118, 0.117, 0.116, 0.111, 0.109, 0.106, 0.110),
    condition = "Heart Disease"
  ),
  tibble(
    year = c(2000, 2004, 2008, 2012, 2016, 2020, 2023),
    rate = c(0.062, 0.070, 0.082, 0.093, 0.105, 0.113, 0.116),
    condition = "Type 2 Diabetes"
  )
)

diet_diseases_trends |>
  ggplot(aes(x = year, y = rate, color = condition)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.8) +
  scale_color_manual(
    values = c(
      "Obesity" = "#e41a1c",
      "Hypertension" = "#ff7f00",
      "Heart Disease" = "#984ea3",
      "Type 2 Diabetes" = "#377eb8"
    )
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.04, 0.45),
    breaks = seq(0.05, 0.45, by = 0.05)
  ) +
  scale_x_continuous(breaks = c(2000, 2004, 2008, 2012, 2016, 2020, 2023)) +
  labs(
    title = "Prevalence of Diet-Related Chronic Diseases Among US Adults (2000–2023)",
    subtitle = "Tracking trends in obesity, high blood pressure, heart disease, and type 2 diabetes",
    x = "Year",
    y = "Prevalence Rate (% of Adults)",
    color = "Condition",
    caption = "Source: CDC National Center for Health Statistics (NHANES & NHIS Reports)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

### Understanding the Diseases

- **Obesity (BMI ≥ 30):** A chronic metabolic condition where excessive body fat accumulates, increasing strain on organs and joints. Diets high in ultra-processed, energy-dense foods and added sugars promote chronic caloric surpluses and weight gain.
- **Hypertension (High Blood Pressure):** A condition where blood flows through arteries at higher-than-normal pressure (consistently ≥ 130/80 mmHg). High dietary sodium, low potassium intake from whole foods, and vascular damage contribute significantly to high blood pressure.
- **Heart Disease:** A group of cardiovascular disorders—primarily coronary artery disease and heart attacks—caused by arterial plaque buildup (atherosclerosis). Unhealthy dietary fats, refined carbs, and chronic inflammation narrow arteries and restrict blood supply to the heart.
- **Type 2 Diabetes:** A metabolic disorder where cells become resistant to insulin, causing blood glucose levels to remain abnormally high. Long-term diets rich in added sugars and ultra-processed carbohydrates are primary drivers of insulin resistance.

:::



