---
title: "NHANES Health Insights"
subtitle: "Exploring Body Mass Index (BMI) trajectories across age and sex"
---

The **National Health and Nutrition Examination Survey (NHANES)** assesses the health and nutritional status of adults and children across the United States.

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(primer.data)
```

```{r}
#| label: bmi-plot
#| fig-width: 10
#| fig-height: 5.5
#| fig-align: center

nhanes |>
  drop_na(age, bmi, sex) |>
  ggplot(aes(x = age, y = bmi, color = sex)) +
  geom_point(alpha = 0.12, size = 1.2) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  facet_wrap(~ sex) +
  scale_color_manual(values = c("Female" = "#d95f02", "Male" = "#1b9e77")) +
  scale_x_continuous(breaks = seq(0, 80, by = 20)) +
  scale_y_continuous(breaks = seq(10, 80, by = 10)) +
  labs(
    title = "Body Mass Index Trajectory Across the Lifespan",
    subtitle = "Comparing BMI trends across age for female and male NHANES participants",
    x = "Age (years)",
    y = "Body Mass Index (kg/m²)",
    caption = "Source: National Health and Nutrition Examination Survey (primer.data::nhanes)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

### Key Observations

- **Childhood to Early Adulthood:** BMI increases sharply from childhood into the twenties as individuals grow.
- **Middle Adulthood:** BMI levels tend to plateau between ages 30 and 60, with the average hovering in the 28–30 kg/m² range.
- **Later Decades:** The trend line stabilizes and gently tapers downward in older age, following comparable paths for both males and females.

## Obesity Rates Over Time

Beyond individual age trajectories, national health surveys track how adult obesity prevalence has shifted across recent decades.

```{r}
#| label: obesity-time-plot
#| fig-width: 9
#| fig-height: 5
#| fig-align: center

obesity_trends <- tibble(
  year = c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020, 2023),
  rate = c(0.305, 0.305, 0.322, 0.343, 0.337, 0.357, 0.349, 0.377, 0.396, 0.424, 0.419, 0.403)
)

obesity_trends |>
  ggplot(aes(x = year, y = rate)) +
  geom_line(color = "#2c7bb6", linewidth = 1.2) +
  geom_point(color = "#2c7bb6", size = 3) +
  geom_text(aes(label = scales::percent(rate, accuracy = 0.1)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.25, 0.48),
    breaks = seq(0.25, 0.45, by = 0.05)
  ) +
  scale_x_continuous(breaks = c(2000, 2004, 2008, 2012, 2016, 2020, 2023)) +
  labs(
    title = "US Adult Obesity Prevalence Over Time (1999–2023)",
    subtitle = "Age-adjusted percentage of adults (aged 20+) with BMI ≥ 30",
    x = "Survey Year",
    y = "Obesity Rate",
    caption = "Source: CDC / National Center for Health Statistics (NHANES)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

- **Rising Trend:** Between 1999–2000 and 2017–2018, the US adult obesity rate increased from 30.5% to over 42%.
- **Recent Leveling:** Rates remained above 41% entering 2020, and the August 2021–August 2023 survey recorded 40.3%, pointing to a recent stabilization.

## Food Processing Trends Over Time (2001–2023)

Research from NHANES dietary recall surveys tracks how the proportion of daily calories coming from ultra-processed foods versus whole or minimally processed foods has shifted among US adults.

```{r}
#| label: upf-time-plot
#| fig-width: 9
#| fig-height: 5
#| fig-align: center

food_trends <- tibble(
  year = c(2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2023),
  upf_rate = c(0.535, 0.543, 0.554, 0.558, 0.561, 0.564, 0.569, 0.567, 0.570, 0.530),
  non_upf_rate = c(0.327, 0.318, 0.305, 0.300, 0.295, 0.291, 0.283, 0.281, 0.274, 0.305)
)

food_trends |>
  ggplot(aes(x = year, y = upf_rate)) +
  geom_line(color = "#e6550d", linewidth = 1.2) +
  geom_point(color = "#e6550d", size = 3) +
  geom_text(aes(label = scales::percent(upf_rate, accuracy = 0.1)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.48, 0.62),
    breaks = seq(0.48, 0.62, by = 0.02)
  ) +
  scale_x_continuous(breaks = c(2002, 2006, 2010, 2014, 2018, 2023)) +
  labs(
    title = "Ultra-Processed Food Consumption Over Time (2001–2023)",
    subtitle = "Percentage of total daily calories from ultra-processed food among US adults",
    x = "Survey Period Ending Year",
    y = "Share of Daily Calories",
    caption = "Source: NHANES dietary intake data / Juul et al. (2022) & NCHS Data Brief No. 536 (2025)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

```{r}
#| label: non-upf-time-plot
#| fig-width: 9
#| fig-height: 5
#| fig-align: center

food_trends |>
  ggplot(aes(x = year, y = non_upf_rate)) +
  geom_line(color = "#2ca25f", linewidth = 1.2) +
  geom_point(color = "#2ca25f", size = 3) +
  geom_text(aes(label = scales::percent(non_upf_rate, accuracy = 0.1)), vjust = -1, size = 3.5) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0.24, 0.36),
    breaks = seq(0.24, 0.36, by = 0.02)
  ) +
  scale_x_continuous(breaks = c(2002, 2006, 2010, 2014, 2018, 2023)) +
  labs(
    title = "Whole & Minimally Processed Food Consumption Over Time (2001–2023)",
    subtitle = "Percentage of total daily calories from unprocessed foods among US adults",
    x = "Survey Period Ending Year",
    y = "Share of Daily Calories",
    caption = "Source: NHANES dietary intake data / Juul et al. (2022) & NCHS Data Brief No. 536 (2025)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 6)),
    plot.subtitle = element_text(color = "#555555", margin = margin(b = 15)),
    plot.caption = element_text(color = "#888888", size = 9, margin = margin(t = 12)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ececec")
  )
```

- **Opposing Trends:** As ultra-processed foods rose from 53.5% in 2001–2002 to 57.0% in 2017–2018, whole and minimally processed foods fell from 32.7% to 27.4%.
- **Recent Shift:** In the August 2021–August 2023 survey cycle, ultra-processed food intake dropped to 53.0%, while whole and minimally processed calories rose back to approximately 30.5%.



